\documentclass[../main/main.tex]{subfiles}

\newdate{date}{15}{10}{2020}


\begin{document}

\marginpar{ \textbf{Lecture 5.} \\  \displaydate{date}. \\ Compiled:  \today.}


RISCRIVERE (ERO ARRIVATA IN RITARDO)



\section{SIRS Model}

The idea is that you assume that you are in the steady state and so all the derivatives goes to zero.

This model allows to interpolate between SIR (\( w=0 \)) and SIS (\( w \rightarrow \infty  \)).

\begin{equation*}
  \dv{s}{t} = \alpha + w r - \beta  s i - \alpha s, \qquad \dv{i}{t} = \beta s i - \mu i - \alpha i, \qquad \dv{r}{t} = \mu i - w r - \alpha r
\end{equation*}
The endemic state can be found by putting the derivatives equal to zero.

In reality people do not become instantaneously infectious, but there is a \textbf{latent period} which is the time between infection and becoming infectious. Indeed, the pathogen replication takes time i.e. viral load too low to transmit the infection.

It is important to remind that the latent period is not the same of the incubation period. An individual can be infectiuous before symptoms. For instance it has a pre-syntomatic infection as in the case of Covid-19.

The simplest idea

\section{SEIR Model}
The equations are the same of before.

If you estimate the time evolution of the SEIR model and SIR model there is a huge difference. For instance we can approximate for the SEIR:
\begin{equation*}
  i_{SEIR} (t) \approx e^{\qty(\sqrt{4(R_0-1)\sigma \mu  + (\sigma + \mu )^2} -(\sigma +\mu ))t/2 } \approx i_0 e^{\qty(\sqrt{R_0} -1 ) \mu t}
\end{equation*}
while for the SIR:
\begin{equation*}
  i_{SIR} (t) \approx i_0 e^{(R_0-1) \mu t}
\end{equation*}

Podromic state means the difference between latent period and ...


\section{}
We solved the SI model analitically and we observe that the growth is as a sigmoid. We have some sort of saturation at 1.

In the SIS things starts to change. We have some sort of treshold phenomena (epidemic treshold).

For the SIR we cannot solve the equation analitically.

We have the SIRS we can interpulate between the two models.

For the SEIR model we have the exposed slowing down the spreading.




\end{document}
